Microlensing as a possible probe of the internal structure of quasars

### 2014-10-05 paper1/discussion.tex
Theory edits

\section{Discussion}

@@

>>>>

\section{Discussion}
[Mihai's contribution]\\
In the current paper we simulate and study the resulting microlensing lightcurves of geometric crescent-shaped sources and compare them with the microlensing lightcurves of other simple mathematically describable source profiles. 
In order to mimic the behaviour of flux of light from the source in the proximity of a fold caustic we make use of the simple approximation described in equation (5). The equation would exhaustively describe the magnification map and offer a good aproximation for the particular microlensing regime that we consider. 
Namely, the shape of the caustic boundary in the proximity of the source in the respective plane can be approximated with a line due to reason that the local radius of curvature of caustic is orders of magnitude greater than the half-light radius of the studied source. 
In particular cases in which the previously mentioned approximation loses its validity the impact on the quality of the lightcurves is not evenly distributed. The shape of the lightcurve will be maintained though the datapoints corresponding to the source position before and during overlapping the caustic will be affected by smaller errors than the datapoints corresponding to later times. \\

The first two source profiles that we consider are the uniform disk and symmetric gaussian source. Both of them can be described by a half-light radius $r_{1/2}$ and a total unlensed light flux $S_0$. With the two parameters constrained the one dimensional profiles as well as the lightcurves of the two source are completely determined, since no free parameter remains. 
The previous statement does not hold for a crescent source. In the case of the crescent source there are in total five parameters: the integrated flux of the source $S_0$, the radii of the bright/dark disk $R_p$/$R_n$ and the displacement of the centers of the two disks on the axes perpendicular and parallel to the caustic $a$ and $b$. 
Two of the parameters can be reduced by expressing the results in terms of $S_0$ and $r_{1/2}$. The later being unique for any set of parameters $R_p, R_n$ and $a^2+b^2$. Moreover, one of the displacemement parameters $b$ has no impact on the one dimensional profile of the source that results from the projection of the source image on an axis perpendicular to the caustic. 
Since the one dimensional source profile that correpsonds to an axis perpendicular to the caustic contains exhaustively all the information regarding the source that can be revealed by the lightcurve, the value of the parameter $b$ does not effect the shape of the lighcurve. 
Nevertheless the $b$ parameter is relevant to the calculation of $r_{1/2}$. It's qualitative effect is to decrease the value of the half-light radius when the absolute value of the parameter is increased.  
With two parameters constrained and another irrelevant to the shape of the lightcurve, two free parameters remain $R_n$ and $a$. Figure 4 reveals that the lighcurve of a crescent source has more visible features than the other two light-curves corresponding to the disc and guasssian shape. There are three regions where $S_1D$ and by inheritance the lighcurve have distinguishable behaviour. The first region that would be recorded on a lightcurve plot represents the preiod of time when the bright disk begins to be overlapped by the caustic and stops when just before the dark disk reaches the caustic. During this period of time the flux of light from the source is increasingly magnified. The second period starts and ends with the overlapping of the dark disc. As a boundary of the two regions there is a distinguishable point where slope of the magnification is drastically changed. This apparent discontinuity in the first derivative of the magnification function is caused by the caustic amplification of the sudden drop in the $S_1d$ function. During the respective period the magnification growth slows or even reverses during the first part of the period and starts to grow faster again as the dark ends its overlap with the caustic. At the point where the dark disc clears the caustic the growth of the magnification is infinite, which appears as a saddle point on the lightcurve. Next, the final period correspond to the case when the dark disc has cleared the caustic and the bright disc continues to overlap with the caustic. During this period the lighcurve reaches a peak that for most of the parameter space is global and for the rest of the parameter space local. At the end of the period the growth of the magnification is negative infinite. the respective point appears as a second saddle point on the lightcurve. Past this point the magnification of any finite source will decay in roughly the same manner. \\

The impact of parameter $a$ on the shape of the lightcurve can be observed in figure 5 (a-var). Sources where the center of the dark disc reaches the caustic before the center of the bright disc are characterized by a smoother broad peak in contrast to the cases where the center of the dark disc reaches the caustic after the brigh disk. In the case of the latter the instance when the dark disc reaches the caustic corresponds to larger and larger magnifications untill it becomes a local and even global peak. The effect of the $R_n$  parameter on the lighcurve is presented in figures 6 and 7. For the particular set of parameters where $R_p = R_n +a$ the third period of time discussed previously does not exist. Particularizing further, if the value of the radius of the dark disc is comparable to the value of the radius of the bright one then the position and shape of the maximum magnification are strikingly different. In case the crescent reaches the caustic with the bright region first, the peak magnification happens when the dark region reaches the caustic and it is characterized by a sharp variation in magnification growth, In the opposite case the peak appears before the end of the bright disc reaches the caustic and shape is smoother. \\

In addition to the previously studied case of a geometric crescent source created from the overlap of a bright and dark disc we have generated the lightcurves of a constant luminosity elliptical crescent and a gaussian elliptical crescent. The first shows striking similarities to the disk crescent source, being characterized by the similar characteristic periods and instances of time. The second, gaussian elliptical crescent, similarly to the symetric gaussian source has a continuous one dimensional flux profile except in the region of the elliptic shadow. The respective region is detected as a period during which the magnification behaves similarly to the corresponding shadow overlap periods of constant brightness sources previously studied. The similarity between the shapes of the lightcurves of the geometric crescent, elliptical crescent and guassian profile suggests that the geometric crescent is a usefull first approximation of the elliptical sources. \\

With the different source profiles and their corresponding lightcurves studied we can change our point of view of the system to that of an observer. The observer would basically detect only the lightcurve of such as source. As described in section ?? the timing of the onset and offset of the previously described periods can be used to estimate the values of the radii and one of the displacement parameters when assuming a geometric crescent, the halfwidths and one displacement parameters when assuming a constant brightness elliptical crescent and the halfwidth of the dark ellipse when assuming an elliptical gaussian crescent. All quantities can be estimated in terms of the relative velocity of the source in a direction perpendicular to the caustic.\\

The previously mentioned abstract parameters can be related to physical quantities specific to the central region of a quasar. Thus the luminous region would correspond to the bright accretion disc that surrounds the black hole. The later's gravity would cause a shadow in the bright region limited by the extent of the event horizon of the black hole. Therefore the radius of the bright disk or halfwidth of the gaussian crescent would provide an estimate of the size of the accretion disk and the radius of the dark disk or halfwidth of the dark ellipse projection would provide and estimate of the Schwarzschild radius of the black hole. Moreover a value of the Schwarzschild radius can be used to determine the mass of the black hole $M \approx \frac{\left(t_3 - t_2 \right) v_p c^2}{2G}$.

The parameters whose values cannot be determined due to the loss of information from the directions parallel to the caustic could be obtained in the eventuality in which the same source crosses multiple caustics that are not parallel. Multiple crossing of caustics can reveal details of the one dimensional flux profile corresponding to multiple distinct directions which would allow the reconstruction of the two dimesnional profile analogous to the process through which a picture of a CT scan is obtained.

In case of a high quality lightcurve with insignificant noise and measurement errors, the parameters can be obtained by simply identifying the characterstic instances of time withought making use of the actual values of the magnification map. If the effect of the errors and the noise distorts the magnification time function enough so that the characteristic points are not identifiable with the characteristic periods still visible, the boundaries of the periods can be roughly estimated. To study if good estimates of the parameters can still be obtained when datapoints are strongly affected by the noise we propose ....

[Irshad's contribution]\\

@@

[Contributions from authors]\\
Prasenjit Saha provided the original idea and plan for the research project as well as multiple contributions to the analysis. Mihai Tomozeiu simulated and studied the ideal behaviour of the microlensing lightcurves for the different source profiles discussed. Joachim Wambsganss provided the numerical code used by Manuel Rabold to create the magnification map and corresponding lighcurves used in the MCMC analysis, analisis that was performed by Irshad Mohammed in the last part of the presented work.

### 2014-10-05 paper1/theory.tex
Theory edits

\begin{align}
 S_{1D}^C &= \frac{2 S_0^C}{\pi \left(R_p^2-R_n^2 \right)} \left\{ \sqrt{R_p^2 - (p-p_{sp})^2}  \Theta \left[ R_p^2 - \left( p-p_{sp} \right)^2 \right] \right.\nonumber\\
 &\qquad \left. {} - \sqrt{R_n^2 - (p-p_{sn})^2 } \Theta \left[ R_n^2 - \left( p-p_{sn} \right)^2 \right] \right\}.
\end{align}

@@

The half-light radius of any source is invariant to any rotational transformation. In the present case of a crescent source the effective radius is dependent on the parameters $R_p$, $R_n$ and $\sqrt{a^2+b^2}$ exclusively. From symmetry considerations the centroid of the source is colinear with the centers of the two disks and it is situated at a distance $d_c$ from the center of the brigth disk. $d_c$ can be computed numerically by solving the equation:

\begin{equation}
\frac{S_0^C}{2} =  
\end{equation}

@@

\begin{figure}
\includegraphics[width = .8\textwidth]{plots/4source_magnification.eps}
\caption{\label{fig:lightcurve_gauss} Lightcurves of crescents, gaussian and disk for sources with identical $S_0$ and $r_{1/2}$. Crescent source has $r_{1/2}$ = 0.72754 $R_p$, $R_n$ = 0.4 $R_p$, a = 0.3/-0.3 $R_p$ }
\end{figure}

@@

\begin{figure}
\includegraphics[width = .8\textwidth]{plots/4avar_magnification.eps}
\caption{\label{fig:lightcurve_disk} Lightcurves of crescent sources with $R_n$ = 0.4 $R_p$.}
\end{figure}

@@

\begin{figure}
\includegraphics[width = .8\textwidth]{plots/4rn_magnification.eps}
\caption{\label{fig:lightcurve_crescent_back} Lightcurves of backwards crescents with  a = 0.3 $R_p$.}
\end{figure}

@@

\begin{figure}
\includegraphics[width = .8\textwidth]{plots/4rn2_magnification.eps}
\caption{\label{fig:lightcurve_crescent} Lightcurve of forwards crescents with a = -0.3 $R_p$.}
\end{figure}

>>>>

\begin{align}
 S_{1D}^C &= \frac{2 S_0^C}{\pi \left(R_p^2-R_n^2 \right)} \left\{ \sqrt{R_p^2 - (p-p_{sp})^2}  \Theta \left[ R_p^2 - \left( p-p_{sp} \right)^2 \right] \right.\nonumber\\
 &\qquad \left. {} - \sqrt{R_n^2 - (p-p_{sn})^2 } \Theta \left[ R_n^2 - \left( p-p_{sn} \right)^2 \right] \right\}.
\label{eqn:s1_d}
\end{align}

@@

The half-light radius of any source is invariant to any rotational transformation. In the present case of a crescent source the effective radius is dependent on the parameters $R_p$, $R_n$ and $\sqrt{a^2+b^2}\equiv c $ exclusively. From symmetry considerations the centroid of the source is colinear with the centers of the two disks and it is situated at a distance $d_c$ from the center of the brigth disk. $d_c$ can be computed numerically with the use of a variation of equation \ref{eqn:s1_d}:

\begin{equation}
\frac{S_0^C}{2} =  \frac{2 S_0^C}{\pi \left(R_p^2-R_n^2 \right)} \int_{d_c}^{R_p} \left[ \sqrt{R_p^2 - p^2} - \sqrt{R_p^2 - \left(p-c\right)^2} \Theta \left(R_p^2 - \left(p-c\right)^2 \right) \right] dp.
\end{equation}

With the position of the centroid determined, the half-light radius can be also be computed numerically:

\begin{equation}
\frac{S_0^C}{2} =  \frac{2 S_0^C}{\pi \left(R_p^2-R_n^2 \right)} \int_{0}^{r_{1/2}} \left[ \Theta_2  - \Theta_1  \right] p dp;
\end{equation}

\begin{equation}
\Theta_1 = \left[ \pi - \arccos \frac{R_p^2 + p^2 -d_c^2}{2 R_p p} - \arccos \frac{R_p^2 - p^2 + d_c^2}{2 R_p d_c} \right] \Theta \left( p - R_p + d_c \right); 
\end{equation}

\begin{equation}
\Theta_2 =  \pi - \arccos \frac{ p^2  + (c+d_c)^2 - R_n^2}{2 p \left(c + d_c \right)} \Theta \left( p + R_n - d_c -c \right). 
\end{equation}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/4source_profile.eps}
	\includegraphics[width = 0.48\textwidth]{plots/4source_magnification.eps}
\caption{\label{fig:lightcurve_gauss} Lightcurves of crescents, gaussian and disk for sources with identical $S_0$ and $r_{1/2}$. Crescent source has $r_{1/2}$ = 0.72754 $R_p$, $R_n$ = 0.4 $R_p$, a = 0.3/-0.3 $R_p$ }
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/4avar_profile.eps}
	\includegraphics[width = 0.48\textwidth]{plots/4avar_magnification.eps}
\caption{\label{fig:lightcurve_disk} Lightcurves of crescent sources with $R_n$ = 0.4 $R_p$.}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/4rn_profile.eps}
	\includegraphics[width = 0.48\textwidth]{plots/4rn_magnification.eps}
\caption{\label{fig:lightcurve_crescent_back} Lightcurves of backwards crescents with  a = 0.3 $R_p$.}
\end{figure}

@@

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/4rn2_profile.eps}
	\includegraphics[width = 0.48\textwidth]{plots/4rn2_magnification.eps}
\caption{\label{fig:lightcurve_crescent} Lightcurve of forwards crescents with a = -0.3 $R_p$.}
\end{figure}

There are four characteristic points visible on the resulting one dimensional light profile of the crescent shaped source. Two of the points, p1 and p4, mark the outer boundaries of the luminous disc component. The other two, p2 and p3, mark the boundaries of the dark disc component. Since $S_1D$ is a projection of $S_2D$ on a line perpendicular to the caustic, the following relation holds:
\begin{equation}
	p_2-p_1 = R_p -R_n - a.
\end{equation}
In addition there are two other obvious relations the characteristic points which are independent of the projection:
\begin{equation}
	p_4 -p_1 = 2 R_p,
\end{equation}

\begin{equation}
        p_3 -p_2 = 2 R_n.
\end{equation}
All four points mark the positions where derivative $\frac{dS_{1D}}{dp}$ is discontinuous. The points can be used to define three regions: $p_1 - p_2$ where $S_1D$ is convex, $p_2 - p_3$ where $S_1D$ is concave, and $p_3 - p_4$ where $S_1D$ is convex again. \\

Due to the nature of the caustic and the monotonic behaviour of the magnification map on both sides of the caustic. , the previously mentioned characteristic points are inherited by the microlensing lightcurve. The points on the temporal dimension $t_1, t_2, t_3$ and $t_4$ corresnpond to instances in time when the fold is aligned with $p_1, p_2, p_3$ and $p_4$, respectively. For a constant relative velocity $v_p$ between the source and the caustic there is a simple relation between the points $p_i$ and instances $t_i$:

\begin{equation}
	t_j - t_i = \frac{p_j - p_i}{v_p}.
\end{equation}

With the use of the previous four equations the following identities can be written:

\begin{equation}
	R_p = \frac{v_p \left( t_4 -t_1 \right)}{2}, 
\end{equation}

\begin{equation}
        R_n = \frac{v_p \left( t_3 -t_2 \right)}{2}, 
\end{equation}

\begin{equation}
        a = \frac{v_p \left( t_4 +t_1 - t_3 - t_2 \right)}{2}. 
\end{equation}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/ch_points.eps}
        \includegraphics[width = 0.48\textwidth]{plots/ch_instances.eps}
\caption{\label{fig:char_points} The characteristic points and instances of a crescent source's one dimensional profile and microlensing lightcurve.}
\end{figure}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/infinite_fold_ellipsess.eps}
\caption{\label{fig:ellipse_profile} Elliptical sources with constant and gaussian distributed brightness.}
\end{figure}

\subsection{Extension to ellipticaly shaped crescent sources}

We further consider a source built from the superposition of two unalligned ellipses of constant surface brightness. Analogous to the previously presented crescent source the dark ellipse is interior to the bright ellipse and the distance between their centers projected on the lines perpendicular and parallel to the fold are $a$ and $b$. The semidimensions are ($a_p, b_p$) for thebright ellipse and ($a_n, b_n$) for the dark ellipse. The semimajor axis of the ellipses for the angles $\delta_p$ and $\delta_n$ with the caustic resulting in projected widths $2 h_p$ and $2h_n$ of the two ellipses on the axis perpendicular to the caustic.

The two dimensional description of the source is:

\begin{equation}
S_{2D}^{CEC} = S_{2D}^{CEP} - S_{2D}^{CEN}
\end{equation}

Where

\begin {equation}
S_{2D}^{CEP} = \frac{S_0^{CEC}}{2 \pi \left(a_p b_p - a_n b_n \right)} \Theta \left( 1 - \frac{\left( \left( p -p_{sp} \right) \sin{\delta_p} + \left( q - q_{sp} \right) \cos{\delta_p} \right)^2}{a_p^2}  - \frac{\left( \left( p -p_{sp} \right) \cos{\delta_p} + \left( q - q_{sp} \right) \sin{\delta_p} \right)^2}{ b_p^2} \right), 
\end {equation}
and
\begin {equation}
S_{2D}^{CEN} = \frac{S_0^{CEC}}{2 \pi \left(a_p b_p - a_n b_n \right)} \Theta \left( 1 - \frac{\left( \left( p -p_{sn} \right) \sin{\delta_n} + \left( q - q_{sn} \right) \cos{\delta_n} \right)^2}{a_n^2}  - \frac{\left( \left( p -p_{sn} \right) \cos{\delta_n} + \left( q - q_{sn} \right) \sin{\delta_n} \right)^2}{ b_n^2} \right), 
\end {equation}.

Analogous to the previous three equations the following relations can be obtained:

\begin{equation}
        h_p = \frac{v_p \left( t_4 -t_1 \right)}{2}, 
\end{equation}

\begin{equation}
        h_n = \frac{v_p \left( t_3 -t_2 \right)}{2}, 
\end{equation}

\begin{equation}
        a = \frac{v_p \left( t_4 +t_1 - t_3 - t_2 \right)}{2}. 
\end{equation}

The time instances used in the previous equations are analogous to the ones considered in the crescent source formed from the superposition of constant surface brightness disks.

Moreover we consider a source described by an elliptical gaussian brightness function shadowed by a dark ellipse. In this case the two dimensional description of the source is given by:

\begin {equation}
S_{2D}^{GEC} = \frac{S_0^{GEC}}{2 \pi a_p b_p} \exp \left( - \frac{\left( \left( p -p_{sp} \right) \sin{\delta_p} + \left( q - q_{sp} \right) \cos{\delta_p} \right)^2}{2 a_p^2}  - \frac{\left( \left( p -p_{sp} \right) \cos{\delta_p} + \left( q - q_{sp} \right) \sin{\delta_p} \right)^2}{2 b_p^2} \right) \Pi \left(p,q\right)
\end {equation}

where the function
\begin{equation}
\Pi \left(p,q\right) = 1 - \Theta \left( 1 - \frac{\left( \left( p -p_{sn} \right) \sin{\delta_n} + \left( q - q_{sn} \right) \cos{\delta_n} \right)^2}{2 a_n^2}  - \frac{\left( \left( p -p_{sn} \right) \cos{\delta_n} + \left( q - q_{sn} \right) \sin{\delta_n} \right)^2}{2 b_n^2} \right). 
\end{equation}

\begin{figure}
\centering
	\includegraphics[width = 0.48\textwidth]{plots/ellipse_profile.eps}
        \includegraphics[width = 0.48\textwidth]{plots/ellipse_lightcurve.eps}
\caption{\label{fig:ellipse_profile} The one dimensional profile and resulting lightcurves of the elliptical sources.}
\end{figure}
